---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
##Introduction

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data was collected for two months (Oct and Nov 2012) from an anonymous individual and include the number of steps taken in 5 minute intervals each day.

The dataset contains 3 variables and 17568 observations over a period of 61 days.

Variables:

steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)

date: The date on which the measurement was taken in YYYY-MM-DD format, there are 61 days in the dataset.

interval: Identifier for the 5-minute interval in which measurement was taken, There are 288 intervals per day (24 hours * 12 intervals of 5 minutes per hour), numbered from 0 to 2355. For each day, the first interval is 0 (00.00pm), the last interval is 2355 (11.55pm)


##Load all required packages
```{r Load Libraries, echo=TRUE, results='hide'}
library(lubridate)
```

## Loading and preprocessing the data
- First the working directory is set.
-The file is then loaded and read

```{r readfile, echo= TRUE}
activity <- read.csv("activity.csv", header = TRUE, na.strings = "NA")

```

-The date is then converted into the right format
```{r, echo = TRUE}
activity$date <- ymd(activity$date)
```

-next we inspect the dataset
```{r summary, echo=}
head (activity)

summary(activity)

str(activity)
```

## What is mean total number of steps taken per day?

- To calculate the mean total number of steps per day, we first use the aggregate function and remove missing values (NA's) as mentioned in the question. 

```{r histogram, echo=TRUE, fig.height=6, fig.width=8}

steps_per_day <- aggregate(steps~date, data = activity, FUN = sum, na.rm=TRUE)

mean_steps <- as.integer(mean(steps_per_day$steps))

median_steps <- as.integer(median(steps_per_day$steps))

hist(steps_per_day$steps, xlab = "Steps per day", ylab = "frequency", 
     main = "Histogram of mean total steps/day", col= "blue")
abline(v=c(mean_steps,median_steps), col=c("black", "red"), lty=c(1,3), lwd=c(1, 3))
```


## What is the average daily activity pattern?

-To calculate the average daily pattern, we will first create a time-series graph with the 5 minute intervals on the x-axis and the average number of steps per day averaged across all days in the y-axis. 

```{r timeseries, echo=TRUE, fig.height=6, fig.width=8}
steps_mean <- aggregate(steps~interval, data = activity, FUN = mean, na.rm=TRUE)


## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
